---
title: "01_Munging"
author: "Jessica Stephens"
date: "7/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Importing Data

data is in a microsoft excel workbook
readxl

```{r load, echo=FALSE}

#check wd and file path match

here("ZMB MWI KEN UGA by age group DataList_7_9_2020 2_37_41 PM.xlsx")
getwd()
setwd("C:/Users/jesse/OneDrive/Documents/Work Material/USAID/Treatment/Treatment_Aging_1yrBands")

#read xlsx - specify path, sheet name to read

df_ken_uga <- read_xlsx("Data/ZMB MWI KEN UGA by age group DataList_7_9_2020 2_37_41 PM.xlsx",
                sheet="KEN and UGA") 


df_zmb_mwi <- read_xlsx("Data/ZMB MWI KEN UGA by age group DataList_7_9_2020 2_37_41 PM.xlsx",
                sheet="ZMB and MWI") 

```

## Check Data Import


```{r load, echo=FALSE}

glimpse(df_ken_uga)
View(df_ken_uga)

```



## Append dataset
We can add both df's together to have one df with all of the observations for the 4 countries as a foundational df
will use bind_rows here in the tidyverse, but union can also be used in base R to see if things are compatible

```{r load, echo=FALSE}

df_combined <- bind_rows(df_ken_uga, df_zmb_mwi)


```

See what that looks like

```{r}
print(df_combined)
view(df_combined)
```








## Clean dataset: Isolate Age from E_Ind and drop unnecessary columns/variables

Goal: the numbers for each one-year age band of people with HIV and of people on ART
# make a new variable with the the values after "Age" and before ";"
```{r load, echo=FALSE}

df_combined_seperate1 <-df_combined  %>% 
    separate(E_Ind, c("A","B", "C", "D", "E"), sep = "([+;])")
view(df_combined_seperate1)

#"B" holds the age


df_combined_seperate2 <- df_combined_seperate1 %>% 
  separate(B, c("X", "Y", "Age_num"), sep = "([ ])")
  view(df_combined_seperate2)

  
df_combined_seperate3<- df_combined_seperate2 %>% 
  separate(A, c("X1", "PLVHIV_or_ART", "X2"), sep = "([-])")
view(df_combined_seperate3)


names(df_combined_seperate3)
#drop 0 from PLHIV vs ART

df_combined_seperate4<- df_combined_seperate3 %>% 
    separate("PLVHIV_or_ART", c("PLVHIV_or_ART", "Y1"), sep = "([0])")
view(df_combined_seperate4)
  
#drop X1, X2, Y, X, C, D, E,Y1
df_combined_seperate5<- df_combined_seperate4 %>% 
    select(E_Count, PLVHIV_or_ART, Age_num, Time, Value, Rounded)
view(df_combined_seperate5)

#rename country variable
df_combined_seperate6<- df_combined_seperate5 %>% 
    rename(Country=E_Count)
view(df_combined_seperate6)
glimpse(df_combined_seperate6)

df_combined_seperate6$Age_num<-
    as.numeric(as.character(unlist(df_combined_seperate6$Age_num)))
view(df_combined_seperate6)
glimpse(df_combined_seperate6)

```


#############################################
# Restructure data
##############################################

## add variables for Age Categories for ACTUAL Age

Create categories for PEPFAR age brackets
<1
1-4
5-9
10-14
15-19
20-24
25-29
30-34
35-39
40-44
45-49
50+

As well as older fine age brackets
<1
1-9
10-14
15-19
20-24
25-29
30-34
35-39
40-44
45-49
50+

https://stackoverflow.com/questions/12979456/r-code-to-categorize-age-into-group-bins-breaks 


```{r load, echo=FALSE}
#find max
range(df_combined_seperate6$Age_num)
#0-80

library(data.table)

PEPFARagebreaks <- c(0,1,5,10,15,20,25,30,35,40,45,50,85)
PEPFARagelabels <- c("<1","1-4","5-9","10-14","15-19","20-24","25-29","30-34",
               "35-39","40-44","45-49","50+")
PEPFARagebreaks18 <- c(0,1,10,15,20,25,30,35,40,45,50,85)
PEPFARagelabels18 <- c("<1","1-9","10-14","15-19","20-24","25-29","30-34",
               "35-39","40-44","45-49","50+")


df_combined_PEPFARage2019 <- setDT(df_combined_seperate6) [ , ACTUAL_Age_cat_Fine := cut(Age_num, 
                                breaks = PEPFARagebreaks, 
                                right = FALSE, 
                                labels = PEPFARagelabels)]
view(df_combined_PEPFARage2019)
glimpse(df_combined_PEPFARage2019)



df_combined_PEPFARage2018 <- setDT(df_combined_PEPFARage2019) [ , ACTUAL_Age_cat_SemiFine := cut(Age_num, 
                                breaks = PEPFARagebreaks18, 
                                right = FALSE, 
                                labels = PEPFARagelabels18)]


view(df_combined_PEPFARage2018)
glimpse(df_combined_PEPFARage2018)
names(df_combined_PEPFARage2018)
#df_combined_PEPFARagegroups <- df_combined_PEPFARage2018 %>%
 # select(!c(agegroups, agegroups18, agegroups19))
#names(df_combined_PEPFARagegroups)
#view(df_combined_PEPFARagegroups)
```

#cant get cut to work below
df_combined_PEPFARage2019 <- df_combined_seperate6 %>% 
    findInterval(df_combined_seperate6$Age_num, c(0,4,9,14,19,24,29,34,39,44,49,80))
    cut(df_combined_seperate6$Age_num, c(0,4,9,14,19,24,29,34,39,44,49,80, right=FALSE))
view(df_combined_PEPFARage2019)
glimpse(df_combined_PEPFARage2019)
###Error in cut.default(df_combined_seperate6$Age_num, c(0, 4, 9, 14, 19, : 'breaks' are not unique








# Figure out predicted ages 
#ex, 2017 ages if 2016 aged w/o other loss/gains


```{r load, echo=FALSE}

df_combined_PREDICTagetime <- df_combined_PEPFARage2018 %>%
 mutate(Age_predict=Age_num+1, Time_predict=Time+1)
view(df_combined_PREDICTagetime)



df_combined_PREDICTagegroups <- setDT(df_combined_PREDICTagetime) [ , PREDICT_Age_cat_Fine := cut(Age_predict, 
                                breaks = PEPFARagebreaks, 
                                right = FALSE, 
                                labels = PEPFARagelabels)]
view(df_combined_PREDICTagegroups)
glimpse(df_combined_PREDICTagegroups)



df_combined_PREDICTagegroups <- setDT(df_combined_PREDICTagegroups) [ , PREDICT_Age_cat_SemiFine := cut(Age_predict, 
                                breaks = PEPFARagebreaks18, 
                                right = FALSE, 
                                labels = PEPFARagelabels18)]


view(df_combined_PREDICTagegroups)
glimpse(df_combined_PREDICTagegroups)
names(df_combined_PREDICTagegroups)




```






## Group by Age Groups and Country 
#actual and predicted separately
#merge actual and predicted

```{r load, echo=FALSE}

#only for semi fine brackets at this time

df_long_Grouped_Actual_SemiFine<- df_combined_PREDICTagegroups %>%
  group_by(Country, PLVHIV_or_ART, Time, ACTUAL_Age_cat_SemiFine) %>%
  summarise(Total_Value_Actual=sum(Value))
glimpse(df_long_Grouped_Actual_SemiFine)
view(df_long_Grouped_Actual_SemiFine)


df_long_Grouped_Predict_SemiFine<- df_combined_PREDICTagegroups %>%
  group_by(Country, PLVHIV_or_ART, Time_predict, PREDICT_Age_cat_SemiFine) %>%
  summarise(Total_Value_Predict=sum(Value))
glimpse(df_long_Grouped_Predict_SemiFine)
view(df_long_Grouped_Predict_SemiFine)


#rename predict variables to match actual so they can merge
df_long_Grouped_Predict_SemiFine_vars<- df_long_Grouped_Predict_SemiFine %>%
  rename(Time=Time_predict, ACTUAL_Age_cat_SemiFine=PREDICT_Age_cat_SemiFine)
glimpse(df_long_Grouped_Predict_SemiFine_vars)
view(df_long_Grouped_Predict_SemiFine_vars)

df_long_Grouped_SemiFine <- df_long_Grouped_Actual_SemiFine %>% 
  left_join(df_long_Grouped_Predict_SemiFine_vars, by=c("Time"="Time", "ACTUAL_Age_cat_SemiFine"="ACTUAL_Age_cat_SemiFine","Country"="Country", "PLVHIV_or_ART"="PLVHIV_or_ART"))
view(df_long_Grouped_SemiFine)





```







## Transform from long to wide

```{r load, echo=FALSE}


df_wide_PEPFARagegroups <- df_long_Grouped_SemiFine %>%
  gather(key="counttype", value= "countvalue", -c(1:4)) %>%
  spread(Time, countvalue) %>%
  mutate(
    counttype=str_remove(counttype,"Total_Value_"),
    Perc_diff_actual= ifelse(counttype=="Actual", (`2019`/`2018`)*100, NA) ) %>%
  #group_by(Country, PLVHIV_or_ART, ACTUAL_Age_cat_SemiFine) %>%
  mutate(
    Perc_diff_perdict= .[counttype == 'Predict',`2019`] / .[counttype == 'Actual',`2018`])
  

    ifelse(!is.na(`2019`) | !is.na(`2018`),

#ex actual 2019 / actual 2018
#predicted 2019/actual 2018

df_wide_PEPFARagegroups <- df_combined_PEPFARage2018 %>%
  pivot_wider(
      names_from=Time,
      names_glue="ACTUAL_Value_{Time}",
      values_from=Value
      )
view(df_wide_PEPFARagegroups)

  

```


## look at difference between acuatal ages - aging in/out actual loss or growth 
#ex actual 2019 / actual 2018

```{r load, echo=FALSE}






mutate(df_long_Grouped_SemiFine
  Actual_Difference= 
    (Total_value_Actual/(Total_Value_Actual/(Time-1)

time-1



  for(Total_value_Actual in df_long_Grouped_SemiFine){
  Actual_Difference <- (Total_value_Actual)/(Total_value_Actual where Time=Time-1)



```



##calc predicted % difference
#predicted 2019/actual 2018



```{r load, echo=FALSE}




  

```


##Compare actual to predicted - expected impact based on aging

```{r load, echo=FALSE}




  

```





## Merge Tx_Curr and TX_new



## Calc predicted remaining in 2019
# prev year tx_curr*prediction % difference




##Calc Expected FY19 CURR accounting for aging in/out
#current FY tx_new*predicted remaning in 2019 (last calculation)

###################################################


## Pull out Uganda

```{r load, echo=FALSE}

df_uga <- df_combined_age_min %>%
  filter(E_Count %in% c("Uganda"))
  

```




## Export CSV

```{r load, echo=FALSE}

write.csv(df_combined_age2,"C:/Users/jesse/OneDrive/Documents/Work Material/USAID/Treatment/Data/ZMB MWI KEN UGA by age group DataList_7_9_2020 2_37_41 PM_AgeSeparated.csv", row.names = FALSE)

write.csv(df_combined_age2_hiv_art,"C:/Users/jesse/OneDrive/Documents/Work Material/USAID/Treatment/Data/Spectrum_ZMB MWI KEN UGA _Age_PLHIV_ART.csv", row.names = FALSE)
#

```




























#################################   ARCHIVE CODE





# Figure out predicted ages 
#ex, 2017 ages if 2016 aged w/o other loss/gains


```{r load, echo=FALSE}



PREDICTagebreaks <- c(0,4,9,14,19,24,29,34,39,44,49,85)
PREDICTagelabels <- c("0-3","4-8","9-13","14-18","19-23","24-28","29-33",
               "34-38","39-43","44-48","49+")

PREDICTagebreaks18 <- c(0,9,14,19,24,29,34,39,44,49,85)
PREDICTagelabels18 <- c("0-8","9-13","14-18","19-23","24-28","29-33",
               "34-38","39-43","44-48","49+")

df_wide_PREDICTagegroupsFINE <- setDT(df_combined_PEPFARagegroups) [ , PREDICT_Age_cat_Fine := cut(Age_num, 
                                breaks = PREDICTagebreaks, 
                                right = FALSE, 
                                labels = PREDICTagelabels)]
view(df_wide_PREDICTagegroupsFINE)
glimpse(df_wide_PREDICTagegroupsFINE)

df_wide_PREDICTagegroupsALL <- setDT(df_wide_PREDICTagegroupsFINE) [ , PREDICT_Age_cat_SemiFine := cut(Age_num, 
                                breaks = PREDICTagebreaks18, 
                                right = FALSE, 
                                labels = PREDICTagelabels18)]
view(df_wide_PREDICTagegroupsALL)
glimpse(df_wide_PREDICTagegroupsALL)

```



## Figure out predicted ages 
#ex, 2017 ages if 2016 aged w/o other loss/gains


```{r load, echo=FALSE}


for each year and actual age cat
  do value of previous year and 


PRED_Value_2019(ACTUAL_Age_cat_Fine) = ACTUAL_Value_2018 (PRED_age_bands)


#loop over age bands

#long

for(i in Value){
  Pred_Value <- when Time=Time-1 & PRED_Age_cat_SemiFine=0


  
 # wide???
    
  sum(ACTUAL_Value_2018) if PRED_age_bands = 
}

PRED_Value_2019_1-9  <- ACTUAL_Value_2018_0-8

PRED_Value_2019_1-9  <- ACTUAL_Value_2018_0-8


```



